# The Normal Distribution and Using Z Scores

## Key Concepts

- Normal Distribution
- Variance, Standard Deviation
- Probability Values
- Cut-Off Points

## Key Formulas

$$z = \frac{X - \mu}{\sigma}$$

$$SE = \frac{\sigma}{\sqrt{n}}$$

$$z = \frac{M - \mu}{\sigma / \sqrt{n}}$$

## Code Examples

```{r}
#| eval: false
library(palmerpenguins)
library(tidyverse)

mu    <- 45.1
sigma <- 5.5

penguins |>
  drop_na(bill_length_mm) |>
  mutate(z = (bill_length_mm - mu) / sigma) |>
  select(species, bill_length_mm, z) |>
  head(8)
```

```{r}
#| eval: false
pnorm(1.0)

z    <- 1.5
mu   <- 45.1
sigma <- 5.5
X    <- z * sigma + mu
X
```

## Exercises

### Conceptual Questions

1. The normal distribution is symmetric about its:

   A. Mode only

   B. Median only

   C. Mean, median, and mode (all three are equal)

   D. Variance

2. Approximately what percentage of scores in a normal distribution fall within ±1 standard deviation of the mean?

   A. 50%

   B. 68%

   C. 95%

   D. 99.7%

3. Approximately what percentage of scores in a normal distribution fall within ±2 standard deviations of the mean?

   A. 50%

   B. 68%

   C. 95%

   D. 99.7%

4. In a one-tailed test where the researcher predicts scores will be *higher* than average, the rejection region is located in:

   A. The lower tail of the distribution

   B. Both tails equally

   C. The upper tail of the distribution

   D. The center of the distribution

5. The null hypothesis for a z-test typically states:

   A. The sample mean is significantly different from the population mean

   B. There is no difference between the sample mean and the population mean

   C. The sample mean is greater than the population mean

   D. The population standard deviation equals zero

6. In a two-tailed test at α = .05, the rejection regions occupy:

   A. The upper 5% of the distribution only

   B. The lower 5% of the distribution only

   C. 2.5% in each tail (5% total)

   D. The central 5% of the distribution

7. The standard error of the mean (SE) is computed as:

   A. $SE = \sigma \cdot \sqrt{n}$

   B. $SE = \dfrac{\sigma}{n}$

   C. $SE = \dfrac{\sigma}{\sqrt{n}}$

   D. $SE = \dfrac{\sigma^2}{n}$

8. When α = .05, the critical z values for a two-tailed test are:

   A. ±1.28

   B. ±1.645

   C. ±1.96

   D. ±2.58

9. TRUE or FALSE: If the obtained z statistic falls beyond the critical value (into the rejection region), we reject the null hypothesis.

   A. True

   B. False

10. TRUE or FALSE: A two-tailed test is appropriate when the researcher has a specific directional prediction (e.g., predicts scores will increase).

    A. True

    B. False

---

### Computational Questions

**Reference Population Parameters**

| Variable | Mean (μ) | SD (σ) |
|:---------|:--------:|:------:|
| Sleep | 8 hours | 1 hour |
| Intelligence | 100 points | 15 points |
| Height | 68 inches | 4 inches |
| Memory | 20 items | 5 items |

```{r}
#| eval: false
populations <- list(
  sleep        = list(mean = 8,   sd = 1),
  intelligence = list(mean = 100, sd = 15),
  height       = list(mean = 68,  sd = 4),
  memory       = list(mean = 20,  sd = 5)
)
```

---

#### Part A: Drawing and Hypothesis Exercises

*For each scenario: (a) draw and label the normal distribution implied by the relevant population parameters, and (b) write the null and alternative hypotheses in both symbolic and plain-language form.*

---

**Exercise 1** — Sleep

Dr. A is interested in whether Patient 47 is sleeping a *different* amount than the population average (in either direction).

a. Draw and label the distribution. *(1 point)*

&nbsp;

&nbsp;

&nbsp;

b. Write H₀ and H₁. *(2 points)*

&nbsp;

&nbsp;

---

**Exercise 2** — Intelligence

Dr. B is interested in whether a new drug (Intelli-Juice) *improves* intelligence scores compared to the known average.

a. Draw and label the distribution. *(1 point)*

&nbsp;

&nbsp;

&nbsp;

b. Write H₀ and H₁. *(2 points)*

&nbsp;

&nbsp;

---

**Exercise 3** — Height

Dr. C is interested in whether a new drug (Shrink-Capsules) *reduces* height compared to the known average.

a. Draw and label the distribution. *(1 point)*

&nbsp;

&nbsp;

&nbsp;

b. Write H₀ and H₁. *(2 points)*

&nbsp;

&nbsp;

---

**Exercise 4** — Memory

Dr. G is interested in whether exercise impacts Gavin's memory in a way that is *different from* the known average (in either direction).

a. Draw and label the distribution. *(1 point)*

&nbsp;

&nbsp;

&nbsp;

b. Write H₀ and H₁. *(2 points)*

&nbsp;

&nbsp;

---

#### Part B: z-Test Decision Problems

*Assume α = .05 unless otherwise noted. Use the population parameters from the table above.*

---

**Problem 5** — Sleep (two-tailed)

A researcher is interested in whether N = 20 patients are getting significantly more or less sleep than average.

a. Is this a one- or two-tailed test? What is/are the cutoff point(s) at α = .05? *(1 point)*

&nbsp;

b. What is/are the cutoff point(s) at α = .10? *(1 point)*

&nbsp;

c. The sample slept an average of 9.8 hours. State your decision at α = .05 and interpret in a complete sentence; report p < .05 or p > .05. *(1 point)*

&nbsp;

&nbsp;

---

**Problem 6** — Intelligence (one-tailed)

A researcher tests whether N = 100 members of a newly discovered alien species are *super intelligent* (i.e., score higher than average).

a. Is this a one- or two-tailed test? What is/are the cutoff point(s) at α = .05? *(1 point)*

&nbsp;

b. What is/are the cutoff point(s) at α = .10? *(1 point)*

&nbsp;

c. The alien sample averages 71 intelligence points. State your decision at α = .05 and interpret; report p < .05 or p > .05. *(1 point)*

&nbsp;

&nbsp;

---

**Problem 7** — Height (one-tailed)

A researcher develops a "shrinking procedure" and measures resulting height for N = 553 participants.

a. Is this a one- or two-tailed test? What is/are the cutoff point(s) at α = .05? *(1 point)*

&nbsp;

b. What is/are the cutoff point(s) at α = .01? *(1 point)*

&nbsp;

c. The shrunken sample has an average height of 48 inches. State your decision at α = .05 and interpret; report p < .05 or p > .05. *(1 point)*

&nbsp;

&nbsp;

---

**Problem 8** — Memory (two-tailed)

A researcher examines whether a new supplement changes memory performance in N = 1,500 volunteers.

a. Is this a one- or two-tailed test? What is/are the cutoff point(s) at α = .05? *(1 point)*

&nbsp;

b. What is/are the cutoff point(s) at α = .01? *(1 point)*

&nbsp;

c. After taking the supplement, the sample recalled an average of 23 items. State your decision at α = .05 and interpret; report p < .05 or p > .05. *(1 point)*

&nbsp;

&nbsp;

---

**Bonus**

What is the formula for a z-test (as opposed to a z-score for a single observation)?

&nbsp;
