# T-Tests: Part #1

## Key Concepts

- Normal Distribution
- Variance, Standard Deviation
- Probability Values
- Cut-Off Points

## Introduction...

## Exercises

### Conceptual Questions

1. Dependent-measures t-tests should be used when:

   A. Whenever possible
   B. We have repeated measurements from the same participants (or matched pairs)
   C. We are interested in a dependent variable
   D. Our independent variable has two levels

2. A t-test is preferred over a z-test when we lack information about the:

   A. Sample size
   B. Population mean
   C. Population standard deviation
   D. Sample variance

3. The t-table provides different cutoff values of t for different:

   A. Variances
   B. Experimental designs
   C. Degrees of freedom (N – 1)
   D. Types of t-tests (independent vs. dependent)

4. TRUE or FALSE: The t distribution looks like the z (normal) distribution, but only when sample size is sufficiently large.

   A. True
   B. False

5. A data set has 100 participants, each contributing one response. The mean of the responses is 75.32. How many responses are free to vary?

   A. 99
   B. 100
   C. 1
   D. Not enough information to know

---

### Computational Questions

*Note: if direction is not specified, use direction implied by hypothesis.*

For each problem: write H₀ and H₁, identify the cutoff value(s), compute the necessary descriptive statistics and t (showing all steps), state your decision, and interpret the result in plain language.

---

**Problem 1** — One-Sample t-Test

An athletic trainer believes a new training protocol increases golf swing speed. The historical population mean is μ₀ = 98 MPH, but the population standard deviation is unknown. N = 12 participants were tested after completing the protocol. Use α = .05.

| Participant | Swing Speed (MPH) |
|:-----------:|:-----------------:|
|  1 | 103 |
|  2 | 102 |
|  3 |  97 |
|  4 | 102 |
|  5 |  99 |
|  6 | 105 |
|  7 |  95 |
|  8 |  98 |
|  9 |  99 |
| 10 | 105 |
| 11 | 101 |
| 12 | 103 |

```{r}
#| eval: false
mu0         <- 98   # historical population mean (MPH)
swing_speed <- c(103, 102, 97, 102, 99, 105, 95, 98, 99, 105, 101, 103)
alpha       <- 0.05
```

Compute the following before finding t:

| Statistic | Value |
|:---|:---|
| Sum = | |
| N = | |
| Mean = | |
| SS = | |
| Variance = | |
| SD = | |
| df = | |

Write H₀/H₁ and determine the cutoff value:

&nbsp;

Compute t, state your decision, and interpret:

&nbsp;

&nbsp;

&nbsp;

---

**Problem 2** — Dependent-Measures (Within-Subjects) t-Test

The same athletic trainer now wants to test whether the protocol increases swing speed *within subjects*. N = 12 participants are tested **before** and **after** the training protocol. Use α = .05.

| Participant | Before (MPH) | After (MPH) | Difference (*D*) | *D* − M̄~D~ | (*D* − M̄~D~)² |
|:-----------:|:------------:|:-----------:|:----------------:|:----------:|:--------------:|
|  1 | 101 | 103 | | | |
|  2 | 102 | 102 | | | |
|  3 |  95 |  97 | | | |
|  4 | 102 | 101 | | | |
|  5 |  99 |  96 | | | |
|  6 | 101 | 105 | | | |
|  7 |  93 |  95 | | | |
|  8 |  97 |  98 | | | |
|  9 |  99 |  98 | | | |
| 10 | 103 | 105 | | | |
| 11 | 101 | 104 | | | |
| 12 | 101 | 102 | | | |
| **Sum** | | | | | |

```{r}
#| eval: false
before <- c(101, 102, 95, 102, 99, 101, 93, 97, 99, 103, 101, 101)
after  <- c(103, 102, 97, 101, 96, 105, 95, 98, 98, 105, 104, 102)
D      <- after - before
alpha  <- 0.05
```

Compute the following:

| Statistic | Value |
|:---|:---|
| Sum of D = | |
| Mean of D = | |
| N = | |
| SS~D~ = | |
| Variance~D~ = | |
| SD~D~ = | |
| df = | |

Write H₀/H₁ and determine the cutoff value:

&nbsp;

Compute t, state your decision, and interpret:

&nbsp;

&nbsp;

&nbsp;

---

**Problem 3** — One-Sample t-Test

You are hired by the Minnesota Vikings to determine if they score a different number of points per game than the NFL average. Use α = .01.

| Parameter | Value |
|:---|:---|
| Sample mean (M) | 28 points |
| Sample SD | 7 |
| Sample size (N) | 16 games |
| Population mean (μ₀) | 24 points |

```{r}
#| eval: false
mu0   <- 24   # NFL average points per game
M     <- 28   # Vikings sample mean
SD    <- 7    # Vikings sample SD
N     <- 16   # number of games
alpha <- 0.01
```

Write H₀/H₁, determine cutoff, compute t, decide, and interpret:

&nbsp;

&nbsp;

&nbsp;

---

**Problem 4** — One-Sample t-Test

A counseling center uses a standardized depression score (lower = fewer symptoms), with a known population mean of μ₀ = 22. Researchers expect the new group-therapy program to decrease scores. Use α = .10.

| Parameter | Value |
|:---|:---|
| Population mean (μ₀) | 22 |
| Sample mean (M) | 19.2 |
| Sample SD | 3.5 |
| Sample size (N) | 36 |

```{r}
#| eval: false
mu0   <- 22    # population mean (lower = fewer symptoms)
M     <- 19.2  # sample mean
SD    <- 3.5   # sample SD
N     <- 36    # sample size
alpha <- 0.10
```

Write H₀/H₁, determine cutoff, compute t, decide, and interpret:

&nbsp;

&nbsp;

&nbsp;
