# T-Tests: Part #2

## Key Concepts

- Independent Samples (Between-Subjects) Design
- Pooled Variance
- Degrees of Freedom
- Standard Error of the Difference

## Key Formulas

$$t = \frac{M_1 - M_2}{s_p\sqrt{\dfrac{1}{n_1} + \dfrac{1}{n_2}}}$$

$$s_p^2 = \frac{SS_1 + SS_2}{df_1 + df_2}$$

$$df = N_{total} - 2$$

## Code Examples

```{r}
#| eval: false
library(palmerpenguins)
library(tidyverse)

two_species <- penguins |>
  filter(species %in% c("Adelie", "Chinstrap")) |>
  drop_na(flipper_length_mm)

t.test(flipper_length_mm ~ species, data = two_species, var.equal = TRUE)
```

```{r}
#| eval: false
penguins |>
  filter(species %in% c("Adelie", "Chinstrap")) |>
  drop_na(flipper_length_mm) |>
  group_by(species) |>
  summarise(M = mean(flipper_length_mm), SD = sd(flipper_length_mm), n = n())
```

## Exercises

### Conceptual Questions

1. In hypothesis testing, the p-value represents:

   A. The probability that the null hypothesis is true

   B. The probability of obtaining a result at least as extreme as the observed data, given H₀ is true

   C. The proportion of participants showing the effect

   D. Effect size

2. As degrees of freedom increase, the shape of the t-distribution:

   A. Becomes wider and flatter

   B. Becomes identical to the chi-square distribution

   C. Becomes narrower and more similar to the normal distribution

   D. Remains unchanged

3. A one-tailed test is appropriate when:

   A. The researcher predicts a specific direction of the effect

   B. No direction is predicted

   C. The sample variance is unknown

   D. The study is underpowered

4. If sample size increases, the standard error will:

   A. Increase

   B. Decrease

   C. Stay the same

   D. Equal the standard deviation

5. If α = .01 instead of .05, what happens?

   A. Type I error risk increases

   B. Type I error risk decreases

   C. Power increases

   D. Power stays the same

6. Statistical significance means:

   A. The result is guaranteed to be practically important

   B. The observed effect, or a more extreme effect, is likely assuming H₀

   C. The observed effect, or a more extreme effect, is unlikely assuming H₀

   D. The sample size is large

7. If an independent-means t-test yields t(18) = 2.45, p = .025 (two-tailed; α = .05), the correct decision is:

   A. Fail to reject H₀

   B. Reject H₀

   C. Accept H₁ at α = .05

   D. Unknown

8. Which of the following assumptions does NOT apply to any kind of t-test?

   A. Observations are independent within each group

   B. Homogeneity of variance across groups

   C. Equal sample sizes

   D. All of the above

9. A one-sample t-test reports df = 374. What was N?

   A. 375

   B. 374

   C. 376

   D. None of the above

---

### Computational Questions

*Note: if direction is not specified, use direction implied by hypothesis.*

For each problem: write H₀ and H₁, identify the cutoff value(s), compute all descriptive statistics and t (showing all steps), state your decision, and interpret the result in plain language.

---

**Problem 1** *(25 points)*

A researcher wants to know whether listening to classical music improves concentration. Eight participants complete a concentration task after sitting in silence for 15 minutes; a separate group of eight complete the same task after listening to 15 minutes of classical music. Higher scores indicate better concentration. Conduct the appropriate test at α = .05.

| Participant | Silence | Participant | Music |
|:-----------:|:-------:|:-----------:|:-----:|
| 1 | 72 | 1 | 80 |
| 2 | 68 | 2 | 78 |
| 3 | 74 | 3 | 82 |
| 4 | 70 | 4 | 75 |
| 5 | 69 | 5 | 79 |
| 6 | 71 | 6 | 77 |
| 7 | 73 | 7 | 83 |
| 8 | 67 | 8 | 81 |

```{r}
#| eval: false
silence <- c(72, 68, 74, 70, 69, 71, 73, 67)
music   <- c(80, 78, 82, 75, 79, 77, 83, 81)
alpha   <- 0.05
```

Compute the following for each group:

| Statistic | Silence | Music |
|:---|:---:|:---:|
| Sum = | | |
| N = | | |
| Mean = | | |
| SS = | | |
| df = | | |

Compute pooled variance, SE, and t:

&nbsp;

&nbsp;

&nbsp;

Write H₀/H₁ and determine the cutoff value:

&nbsp;

State your decision and interpret:

&nbsp;

&nbsp;

---

**Problem 2** *(5 points)*

A researcher compares reaction times between caffeine and placebo groups, expecting a reduction in reaction times following caffeine consumption. Name and conduct the appropriate t-test at α = .01. State hypotheses, show all steps, and interpret.

| Parameter | Caffeine | Placebo |
|:---|:---:|:---:|
| N | 12 | 12 |
| Mean (M) | 248 ms | 262 ms |
| SD | 15 | 18 |

```{r}
#| eval: false
M_caffeine <- 248; SD_caffeine <- 15; n_caffeine <- 12
M_placebo  <- 262; SD_placebo  <- 18; n_placebo  <- 12
alpha      <- 0.01
```

Compute SS, pooled variance, SE, df, and t:

&nbsp;

&nbsp;

&nbsp;

Write H₀/H₁, determine cutoff, state decision, and interpret:

&nbsp;

&nbsp;
