# One-Way ANOVA

## Key Concepts

- Normal Distribution
- Variance, Standard Deviation
- Probability Values
- Cut-Off Points

## Key Formulas

$$F = \frac{MS_{between}}{MS_{within}}$$

$$MS_{between} = \frac{SS_{between}}{df_{between}}, \qquad MS_{within} = \frac{SS_{within}}{df_{within}}$$

$$df_{between} = k - 1, \qquad df_{within} = N - k$$

## Code Examples

```{r}
#| eval: false
library(palmerpenguins)
library(tidyverse)

penguins |>
  drop_na(flipper_length_mm) |>
  group_by(species) |>
  summarise(M = mean(flipper_length_mm), SD = sd(flipper_length_mm), n = n())
```

```{r}
#| eval: false
penguins_clean <- penguins |> drop_na(flipper_length_mm)

aov(flipper_length_mm ~ species, data = penguins_clean) |> summary()
```

## Exercises

### Conceptual Questions

1. The main purpose of a one-way ANOVA is to:

   A. Test whether the variance of one sample equals another sample

   B. Test whether three or more group means differ significantly, using variance information

   C. Compare exactly two means

   D. Measure the correlation between two variables

2. In a one-way ANOVA, the total variability in the data is divided into:

   A. Within-group and between-group variance components

   B. Experimental error and measurement bias

   C. Main effects and interactions

   D. Population variance and sampling variance

3. You are comparing exam performance across three different study methods. Which test is most appropriate?

   A. One-sample t-test

   B. Independent-means t-test

   C. One-way ANOVA

   D. Chi-square test

4. Why is it problematic to run multiple t-tests instead of an ANOVA when comparing 3 or more groups?

   A. t-tests cannot handle unequal variances

   B. t-tests reduce statistical power

   C. Multiple t-tests inflate the Type I error rate

   D. ANOVA is easier to compute

5. The correct null hypothesis for a one-way ANOVA with three groups is:

   A. At least one group mean differs

   B. All group means are equal

   C. The variances across groups are equal

   D. Each sample mean equals zero

6. The F statistic in ANOVA is always positive because:

   A. Group means cannot differ in a negative direction

   B. The denominator in F is always negative

   C. Negative F values are omitted by convention

   D. F is a ratio of squared quantities (variances)

7. Under the null hypothesis, the expected value of the F ratio is approximately:

   A. 0

   B. 1

   C. −1

   D. None of the above

8. An ANOVA yields F(2, 27) = 5.41, p = .010. Assuming α = .05, what should you conclude?

   A. At least one group mean differs — reject the null

   B. All group means are equal — fail to reject the null

   C. Not enough information to decide

   D. Each sample mean equals zero

9. An ANOVA yields F(2, 27) = 5.41, p = .010. How many **groups** are being compared?

   A. 3

   B. 4

   C. 2

   D. Not enough information to decide

10. An ANOVA yields F(2, 27) = 5.41, p = .010. How many **participants** were included in this analysis?

    A. 3

    B. 27

    C. 30

    D. Not enough information to decide

---

### Computational Questions

*Note: if direction is not specified, use direction implied by hypothesis.*

For each problem:

- Write out the null and alternative hypotheses
- Find and report the critical/cutoff F value
- Compute all intermediate values, showing steps (group means, SS between, SS within, df between, df within, MS between, MS within, F)
- Complete an ANOVA summary table
- Report the result: F(df~between~, df~within~) = F-value, p < or > α
- Interpret the result in plain language

---

**Problem 1**

A researcher tests how puzzle solve-time (minutes) changes when people listen to different types of music. Fifteen participants were randomized into three conditions (n = 5 each). Use α = .05.

| Participant | No Music | Classical | Popular |
|:-----------:|:--------:|:---------:|:-------:|
| 1 | 12 | 15 | 10 |
| 2 | 10 | 14 | 12 |
| 3 |  9 | 13 | 11 |
| 4 | 11 | 16 |  9 |
| 5 |  8 | 15 | 10 |

```{r}
#| eval: false
no_music  <- c(12, 10, 9, 11, 8)
classical <- c(15, 14, 13, 16, 15)
popular   <- c(10, 12, 11, 9, 10)
alpha     <- 0.05
```

ANOVA Summary Table:

| Source | SS | df | MS | F |
|:---|:--:|:--:|:--:|:--:|
| Between | | | | |
| Within  | | | | |
| **Total** | | | | |

&nbsp;

&nbsp;

&nbsp;

---

**Problem 2**

A researcher has partially completed their analysis. Use the information below to complete the ANOVA table, make a decision at α = .05, and provide all requested interpretations.

| Source | SS | df | MS | F |
|:---|:--:|:--:|:--:|:--:|
| Between | 240 | 2 | | |
| Within  | 240 | 12 | | |
| **Total** | | | | |

&nbsp;

&nbsp;

&nbsp;
